---
title: "Public Bike Share Data Analysis"
author: "A20495939 - Jasleen Bhatia"
author: "A20493607 - Bharath Bandaru"
author: "A20491089  - Himamshu Lakkaraju"
--- 

# Loading Libraries #

```{r}
install.packages("caret", repos = "http://cran.us.r-project.org")
library(caret)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(sqldf)

```

# Loading Dataset #

```{r}
divyydata <- read.csv("Divvy_processed.csv",sep=",",header=TRUE)
# TODO summary(divyydata)
str(divyydata)
# converting data to numeric
divyydata$trip_duration <- as.numeric(divyydata$trip_duration)
colSums(is.na(divyydata))
# removing columns containing NA's
divyydata <- divyydata[complete.cases(divyydata), ] 
divyydata <- divyydata[!(divyydata$trip_duration > 20000),]
# splitting date and creating year
years <- str_split_fixed(divyydata$k_date, "-", 3)
divyydata$year <- as.factor(years[,3])
rm(years)
#converting columns to factors
as.factor.cols <- c("start_station_name", "end_station_name", "k_time",
          "time_of_day","season","is_holiday","day_of_week")
divyydata[as.factor.cols] <- lapply(divyydata[as.factor.cols], factor)

```

```{r}
smp_size <- floor(0.001 * nrow(divyydata))
set.seed(123)
train_ind <- sample(seq_len(nrow(divyydata)), size = smp_size)

new_train <- divyydata[train_ind, ]
new_test <- divyydata[-train_ind, ]

divyydata = new_train

summary(divyydata)
divyydata[1,]
is.na(divyydata)
```
### Data Preprocessing ###
Performing Data Sanity Checks before proceeding with analysis

```{r}
divyydata["Month"]<-format(as.Date(divyydata$k_date,format="%m-%d-%y"),"%m")
head(divyydata,10)
#Checking the shape of the dataset
row=nrow(divyydata)
col=ncol(divyydata)
sprintf("The rows and colums are: %s %s",row,col)
```

```{r}
#See whether missing values or not
sapply(divyydata, function(x) sum(is.na(x)))
```


### Exploratory Data Analysis ###
# Number of users over the years #

```{r}
divyydata$Month <- factor(divyydata$Month)
ggplot()+
  geom_bar(data=divyydata,mapping=aes(x=Month))+
  scale_y_continuous(breaks = seq(0,1000000,100000),labels=scales::comma)+
  labs(x="Month",
       y="Number of Users")+
  labs(title="Number of users over the years")+
  theme(plot.title =element_text(hjust = 0.5,size=15), 
        legend.position = c(.50, .50),
        legend.justification = c("right", "top"),
       axis.title = element_text(size=12),
        axis.text = element_text(size=09))+
  theme(plot.caption=element_text(size=10))
```
# Rideable type over the years #

```{r}
ggplot(data=divyydata)+geom_bar(mapping = aes(x=Month,fill=rideable_type))+
  scale_y_continuous(breaks = seq(0,700000,100000),labels=scales::comma)+
  labs(x="Month",
       y="Number of rideable type",
       fill="Rideable Type")+
  labs(title="Change of usage on rideable types over the years",subtitle = "3 Types of Ride: Classic, Docked, Electric")+
  theme(plot.title =element_text(hjust = 0.5,size=15), 
        legend.position = c(.50, .50),
        legend.justification = c("right", "top"),
       axis.title = element_text(size=12),
        axis.text = element_text(size=09))+
  theme(plot.caption=element_text(size=10))
```

# Ride Length over the years #

```{r}

divyydata["ride_length"] = difftime(divyydata$ended_at,divyydata$started_at, units="mins")
df<-sqldf("select avg(ride_length) as avg_ride,Month from divyydata group by Month")
df$Month <- factor(df$Month)
ggplot(data = df)+
  geom_line(mapping = aes(x=Month,y=avg_ride,group = 1))+
  labs(x="Month",y="Average ride length per minute")+
  labs(title="Change of average ride length over the years")+
    theme(plot.title =element_text(hjust = 0.5,size=15), 
        legend.position = c(.50, .50),
        legend.justification = c("right", "top"),
       axis.title = element_text(size=12),
        axis.text = element_text(size=09))+
  theme(plot.caption=element_text(size=10))
```
#Top 20 most Popular Stations #

```{r}
trip_df<-sqldf("select start_station_name,end_station_name,(start_station_name|| ' to ' ||end_station_name) as station_name,count(ride_id) as number_of_trips from divyydata group by start_station_name,end_station_name order by count(ride_id) desc LIMIT 20")
ggplot(data=trip_df,aes(x=number_of_trips, y=station_name))+geom_bar(stat='identity')+ 
  labs(x="Number of Trips",y="Statiocn Name")+
  labs(title=" Top 20 Popular Stations ")+
  theme(plot.title =element_text(hjust = 0.5,size=15), 
        legend.position = c(.50, .50),
        legend.justification = c("right", "top"),
       axis.title = element_text(size=12),
        axis.text = element_text(size=09))+
  theme(plot.caption=element_text(size=10))
```


```{r}
divyydata_combined <- read.csv("/Users/jasleenkaurbhatia/Downloads/day.csv",sep=",",header=TRUE,quote="",stringsAsFactors=FALSE)
divyydata_combined$actual_temp <- divyydata_combined$temp*41
divyydata_combined$actual_feel_temp <- divyydata_combined$atemp*50
divyydata_combined$actual_windspeed <- divyydata_combined$windspeed*67
divyydata_combined$actual_humidity <- divyydata_combined$hum*100
divyydata_combined$mean_acttemp_feeltemp <- (divyydata_combined$actual_temp+divyydata_combined$actual_feel_temp)/2
str(divyydata_combined)
h <- hist(divyydata_combined$cnt, breaks = 25, ylab = 'Frequency of Rental', xlab = 'Total Bike Rental Count', main = 'Distribution of Total Bike Rental Count', col = 'blue' )
xfit <- seq(min(divyydata_combined$cnt),max(divyydata_combined$cnt), length = 50)
yfit <- dnorm(xfit, mean =mean(divyydata_combined$cnt),sd=sd(divyydata_combined$cnt))
yfit <- yfit*diff(h$mids[1:2])*length(divyydata_combined$cnt)
lines(xfit,yfit, col='red', lwd= 3)
```

#### MAIN DATASET  PLOTS
```{r}
bk_ls = aggregate(x= divyydata$year,by= list(divyydata$bike_id),FUN=sum)

## TRIP COUNTS

#bike id graph
bike_trips = divyydata %>% count(year,bike_id)
yearList = unique(bike_trips$year)
bike_trips = bike_trips[order(bike_trips$year,bike_trips$n, decreasing = T),]

trip_df <- data.frame(filter(bike_trips,bike_trips$year==13)[1:5,])
for(val in yearList[2:7]){
  trip_df = rbind(trip_df,filter(bike_trips,bike_trips$year==val)[1:5,])
}
trip_df

ggplot(data = trip_df, aes(x = n, y= bike_id, fill=year) )+
  geom_bar(stat = "identity")

rm(bike_trips)

# holiday no of trips graph
holiday_trip = divyydata %>% count(is_holiday,year)

ggplot(data = holiday_trip, aes(x = year, y= n, fill=is_holiday) )+
  geom_bar(stat = "identity")

# holiday trip meaned graph
holiday_trip_du = divyydata %>% count(is_holiday,k_date,year)

meaned_holi = holiday_trip_du %>%
    group_by(year,is_holiday) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_holi, aes(x = year, y= Mean, fill=is_holiday) )+
  geom_bar(stat = "identity",position = position_dodge())

# weekdays trip meaned graph
weekday_df = divyydata %>% count(day_of_week,k_date,year)

meaned_wd = weekday_df %>%
    group_by(year,day_of_week) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_wd, aes(x = year, y= Mean, fill=day_of_week) )+
  geom_bar(stat = "identity",position = position_dodge())

# season graph
season_df = divyydata %>% count(season,k_date,year)

meaned_season = season_df %>%
    group_by(year,season) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_season, aes(x = year, y= Mean, fill=season) )+
  geom_bar(stat = "identity",position = position_dodge())

# time graph
time_df = divyydata %>% count(k_time,k_date,year)

meaned_time = time_df %>%
    group_by(year,k_time) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_time, aes(x = year, y= Mean, fill=k_time) )+
  geom_bar(stat = "identity",position = position_dodge())

## TRIP DURATION
start_st = divyydata %>% count(start_station_name,year)


```

#### MAIN GROUPED DATASETS
```{r}
#load data
library(ggplot2)
library(caret)
library(sqldf)
options(scipen = 999)

start_df <-  read.csv("start_date_grouped_fin.csv",header=T)
end_df <-  read.csv("end_date_grouped_fin.csv",header=T)
summary(start_df)
str(end_df)

# converting columns to factors
as.factor.cols <- c("is_holiday", "day_of_week", "season_num", "year","day","month")
start_df[c("start_station_name",as.factor.cols)] <- lapply(start_df[c("start_station_name",as.factor.cols)], factor)
end_df[c("end_station_name",as.factor.cols)] <- lapply(end_df[c("end_station_name",as.factor.cols)], factor)

colSums(is.na(start_df))
colSums(is.na(end_df))
```

#### Plots
```{r}
# plot top stations based on START station
start_top_stations = aggregate(x= start_df$num_of_trips,
	          by= list(start_df$start_station_name),
	          FUN=sum)
start_top_stations = start_top_stations[order(start_top_stations$x,decreasing=T),]
start_top_stations = start_top_stations[1:10,]

ggplot(data=start_top_stations, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))

rm(start_top_stations)
# plot top stations based on END station
end_top_stations = aggregate(x= end_df$count,
	          by= list(end_df$end_station_name),
	          FUN=sum)
end_top_stations = end_top_stations[order(end_top_stations$x,decreasing=T),]
end_top_stations = end_top_stations[1:10,]

ggplot(data=end_top_stations, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))
rm(end_top_stations)

holiday = aggregate(x= end_df$count,
	          by= list(end_df$is_holiday,end_df$end_station_name),
	          FUN=sum)
```


```{r}
lm_test<- lm(divyydata_combined$cnt~divyydata_combined$actual_temp)
summary(lm_test)
plot(lm_test, col = "green")

```



### Data Modelling ###


# DATA 1
#### Loading pre processed Data and factoring required columns 
#### Split data to train and test 
```{r}
head(divyydata)
set.seed(13)

smp_size <- floor(0.01 * nrow(divyydata))
set.seed(123)
train_ind <- sample(seq_len(nrow(divyydata)), size = smp_size)

train <- divyydata[train_ind, ]
test <- divyydata[-train_ind, ]

summary(train)

```

#### Training model with glm and check relevant predictors
```{r}
## lasso to eliminate the predicates
library(fastDummies)
library(glmnet)
sum(is.na(train$bike_id))
train <- train[complete.cases(train), ]
unique(train$season)
train_ex <- dummy_cols(train, select_columns = c("time_of_day", "season","is_holiday","day_of_week","year"), remove_selected_columns =TRUE)
str(train_ex)

# lasso inputs
train_ex[c(10:46)]
train_ex[c(10:46)] <- scale(train_ex[c(10:46)],center = F,scale = T)

x <- data.matrix(train_ex[c(10:46)])
y <- train_ex$trip_duration
lasso_model <- cv.glmnet(x, y, alpha = 1)

best_lambda <- lasso_model$lambda.min
best_lambda
plot(lasso_model) 

library(corrplot)
cor(train_ex[c(10:30)])
corrplot(cor(train_ex[c(10:30)]))
best_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(best_model)

```

#### GLM Model
```{r}
str(train)
smp_size <- floor(0.1 * nrow(train))
set.seed(123)
train_ind <- sample(seq_len(nrow(train)), size = smp_size)

new_train <- train[train_ind, ]
new_test <- train[-train_ind, ]
# GLM model
lmodel_fit <- glm(trip_duration~bike_id+trip_duration+start_station_name+
                    end_station_name+maxtempC+mintempC+totalSnow_cm+uvIndex+
                    h.tempC+h.windSpeedKmph+h.precipMM+h.humidity+h.visibilityKM+
                    h.cloudcover+day_of_week+time_of_day+season+is_holiday,
                  data = new_train)
summary(lmodel_fit)
plot(lmodel_fit)
confint(lmodel_fit)
pred <- predict(lmodel_fit, test)


```

# DATA 2
```{r}
# *** duplicate code for loading data
library(ggplot2)
library(caret)
library(sqldf)
options(scipen = 999)

start_df <-  read.csv("start_date_grouped_fin.csv",header=T)
end_df <-  read.csv("end_date_grouped_fin.csv",header=T)
summary(start_df)
str(end_df)

# converting columns to factors
as.factor.cols <- c("is_holiday", "day_of_week", "season_num", "year","day","month")
start_df[c("start_station_name",as.factor.cols)] <- lapply(start_df[c("start_station_name",as.factor.cols)], factor)
end_df[c("end_station_name",as.factor.cols)] <- lapply(end_df[c("end_station_name",as.factor.cols)], factor)
# *** end of duplicate code for loading data


start_df$is_holiday = as.numeric(start_df$is_holiday)
start_df$day_of_week = as.numeric(start_df$day_of_week)
start_df$season_num = as.numeric(start_df$season_num)
start_df$year = as.numeric(start_df$year)
summary(start_df)
#tt <- as.numeric(tt)
pca2= prcomp(start_df[c(3:18)], scale = TRUE, center = TRUE) 
summary(pca2)
screeplot(pca2,type = 'lines')
biplot(pca2)



#split <- createDataPartition(c(start_df$num_of_trips,start_df$start_station_name,start_df$day,start_df$year,start_df$season_num), p = 0.01, list = F)
#training set
#train <- start_df[split,]  
#test set
#test <- start_df[-split,]
# GLM model
lmodel_fit <- glm(num_of_trips~uvIndex+tempC+totalSnow_cm+
                   mintempC +maxtempC+windSpeedKmph+precipMM+humidity+visibilityKM+
                    cloudcover+is_holiday+day_of_week+season_num+start_station_name,data = train)
summary(lmodel_fit)
plot(lmodel_fit)
confint(lmodel_fit)

options(scipen = 999)
predOut <- predict(lmodel_fit, test, type = "response")
print(predOut)
summary(train)
sum(is.na(start_df$num_of_trips))
smp_size <- floor(0.05 * nrow(start_df))
split <- sample(seq_len(nrow(start_df)), size = smp_size)

train <- start_df[split, ]
test <- start_df[-split, ]
## RF
library(randomForest)

rf <- randomForest(
  num_of_trips ~ start_station_name+totalSnow_cm+tempC+windSpeedKmph+precipMM+visibilityKM+year+is_holiday+day_of_week,
  data=train,na.action=na.exclude
)
head(test)$num_of_trips
pred = predict(rf, newdata=head(test))
pred
```

### Data Visualization ##
```{r}

```


### Model Evaluation ###

```{r}

```


```{r}
rm(divyydata)
```
