---
title: "Public Bike Share Data Analysis"
author: "A20495939 - Jasleen Bhatia, A20491089  - Himamshu Lakkaraju, A20493607 - Bharath Bandaru"
output: html_document
--- 

# Loading Libraries #

```{r}
install.packages("caret", repos = "http://cran.us.r-project.org")
library(caret)
library(dplyr)
library(stringr)
library(ggplot2)
library(sqldf)

```

# Loading Dataset #

```{r}
divyydata <- read.csv("Divvy_processed.csv",sep=",",header=TRUE)
summary(divyydata)
str(divyydata)
# converting data to numeric
divyydata$trip_duration <- as.numeric(divyydata$trip_duration)
# removing columns containing NA's
divyydata <- divyydata[complete.cases(divyydata), ] 

divyydata <- divyydata[!(divyydata$trip_duration > 20000),]
# splitting date and creating year
years <- str_split_fixed(divyydata$k_date, "-", 3)
divyydata$year <- as.factor(years[,3])

unique(divyydata$year)
rm(years)
#converting columns to factors
as.factor.cols <- c("start_station_name", "end_station_name", "k_time",
          "time_of_day","season","is_holiday","day_of_week")
divyydata[as.factor.cols] <- lapply(divyydata[as.factor.cols], factor)
summary(divyydata)

```

### Data Preprocessing ###
Performing Data Sanity Checks before proceeding with analysis
```{r}
divyydata["Month"]<-format(as.Date(divyydata$k_date,format="%m-%d-%y"),"%m")
summary(divyydata)
head(divyydata,10)
#Checking the shape of the dataset
row=nrow(divyydata)
col=ncol(divyydata)
sprintf("The rows and colums are: %s %s",row,col)
```

```{r}
#See whether missing values or not
sapply(divyydata, function(x) sum(is.na(x)))
```


### Exploratory Data Analysis ###

# Ride Length over the years #
```{r}

df<-sqldf("select avg(trip_duration) as avg_ride,year from divyydata group by year")
ggplot(data = df)+
  geom_line(mapping = aes(x=year,y=avg_ride,group = 1))+
  labs(x="year",y="Average ride length per seconds")+
  labs(title="Change of average ride length over the years")+
    theme(plot.title =element_text(hjust = 0.5,size=15), 
        legend.position = c(.50, .50),
        legend.justification = c("right", "top"),
       axis.title = element_text(size=12),
        axis.text = element_text(size=09))+
  theme(plot.caption=element_text(size=10))
```

#### OLD DATASET  PLOTS
```{r}
old_divvy <- read.csv("processed_fin.csv",sep=",",header=TRUE)

years <- str_split_fixed(old_divvy$k_date, "-", 3)
old_divvy$year <- as.factor(years[,3])

unique(old_divvy$year)
rm(years)

old_divvy$bike_id = as.factor(old_divvy$bike_id)

cols <- c("year", "bike_id")
old_divvy = old_divvy[cols]

```

# most used bikes per year #
```{r}
#bike id graph
bike_trips = old_divvy[cols] %>% count(year,bike_id)
yearList = unique(bike_trips$year)
bike_trips = bike_trips[order(bike_trips$year,bike_trips$n, decreasing = T),]

trip_df <- data.frame(filter(bike_trips,bike_trips$year==13)[1:5,])
for(val in yearList[2:7]){
  trip_df = rbind(trip_df,filter(bike_trips,bike_trips$year==val)[1:5,])
}
trip_df

ggplot(data = trip_df, aes(x = n, y= bike_id, fill=year) )+
  geom_bar(stat = "identity")

rm(bike_trips)
rm(old_divvy)
```

# Number of trips per year #
```{r}
# holiday no of trips graph
holiday_trip = divyydata %>% count(is_holiday,year)

ggplot(data = holiday_trip, aes(x = year, y= n, fill=is_holiday) )+
  geom_bar(stat = "identity")+
  labs(x="year",
       y="Number of trips")

# holiday trip meaned graph
holiday_trip_du = divyydata %>% count(is_holiday,k_date,year)

meaned_holi = holiday_trip_du %>%
    group_by(year,is_holiday) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_holi, aes(x = year, y= Mean, fill=is_holiday) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="year",
       y="Average number of trips per day")

```

# Average Number of trips(days) per day over year #
```{r}
# weekdays trip meaned graph
weekday_df = divyydata %>% count(day_of_week,k_date,year)

meaned_wd = weekday_df %>%
    group_by(year,day_of_week) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_wd, aes(x = year, y= Mean, fill=day_of_week) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="year",
       y="Average number of trips per day")
```

# Average Number of trips(season) per day over year #
```{r}
# season graph
season_df = divyydata %>% count(season,k_date,year)

meaned_season = season_df %>%
    group_by(year,season) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_season, aes(x = year, y= Mean, fill=season) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="year",
       y="Average number of trips per day")
```

# Number of trips around time over years #
```{r}
# time graph
time_df = divyydata %>% count(k_time,k_date,year)

meaned_time = time_df %>%
    group_by(year,k_time) %>%
    dplyr::summarize(Mean = mean(n, na.rm=TRUE))

ggplot(data = meaned_time, aes(x = year, y= Mean, fill=k_time) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="year",
       y="Average number of trips around k_time")

```



# Trip duration over years #
```{r}
# time graph
trip_du = divyydata %>%
    group_by(year) %>%
    dplyr::summarize(Mean = mean(trip_duration, na.rm=TRUE))

ggplot(data = trip_du, aes(x = year, y= Mean) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="year",
       y="Average Trip Duration(seconds)")

```

# Trip duration over seasons #
```{r}
# time graph
trip_du = divyydata %>%
    group_by(season) %>%
    dplyr::summarize(Mean = mean(trip_duration, na.rm=TRUE))

ggplot(data = trip_du, aes(x = season, y= Mean ) )+
  geom_bar(stat = "identity",position = position_dodge())+
  labs(x="season(all years)",
       y="Average Trip Duration(seconds)")

```


#### NUMBER OF TRIPS GROUPED DATASETS
```{r}
#load data
library(ggplot2)
library(caret)
library(sqldf)
options(scipen = 999)

start_df <-  read.csv("start_date_grouped_fin.csv",header=T)
end_df <-  read.csv("end_date_grouped_fin.csv",header=T)
str(end_df)

# converting columns to factors
as.factor.cols <- c("is_holiday", "day_of_week", "season_num", "year","day","month")
start_df[c("start_station_name",as.factor.cols)] <- lapply(start_df[c("start_station_name",as.factor.cols)], factor)
end_df[c("end_station_name",as.factor.cols)] <- lapply(end_df[c("end_station_name",as.factor.cols)], factor)
summary(start_df)
colSums(is.na(start_df))
colSums(is.na(end_df))
```

# overall weather plots #
```{r}



par(mfrow = c(2, 1))
plot(x=start_df$tempC,y=start_df$count,xlab = "Temperature",sub = "Temperature vs no. of trip",ylab = "Number of trips(all data)")
plot(x=start_df$totalSnow_cm,y=start_df$count,xlab = "Snow",sub = "Snow vs no. of trip",ylab = "Number of trips(all data)")


par(mfrow = c(2, 1))
plot(x=start_df$windSpeedKmph,y=start_df$count,xlab = "Wind",sub = "Wind vs no. of trip",ylab = "Number of trips(all data)")
plot(x=start_df$cloudcover,y=start_df$count,xlab = "Cloud",sub = "Cloud vs no. of trip",ylab = "Number of trips(all data)")


plot(x=start_df$precipMM,y=start_df$count,xlab = "Rain",sub = "Rain vs no. of trip",ylab = "Number of trips(all data)")

par(mfrow = c(2, 1))
library(dplyr)
meaned_wd = start_df %>%
    group_by(k_date) %>%
    dplyr::summarize(num_trips = mean(count, na.rm=TRUE),temp = mean(tempC, na.rm=TRUE))

plot(x=meaned_wd$temp,y=meaned_wd$num_trips,xlab = "Temperature",sub = "Temperature vs no. of trip",ylab = "Number of trips(per day)")

meaned_wd = start_df %>%
    group_by(k_date) %>%
    dplyr::summarize(num_trips = mean(count, na.rm=TRUE),cloud = mean(cloudcover, na.rm=TRUE))

plot(x=meaned_wd$cloud,y=meaned_wd$num_trips,xlab = "Cloud",sub = "Cloud vs no. of trip",ylab = "Number of trips(per day)")

par(mfrow = c(2, 1))
meaned_wd = start_df %>%
    group_by(k_date) %>%
    dplyr::summarize(num_trips = mean(count, na.rm=TRUE),wind = mean(windSpeedKmph, na.rm=TRUE))

plot(x=meaned_wd$wind,y=meaned_wd$num_trips,xlab = "Wind",sub = "Wind vs no. of trip",ylab = "Number of trips(per day)")


meaned_wd = start_df %>%
    group_by(k_date) %>%
    dplyr::summarize(num_trips = mean(count, na.rm=TRUE),snow = mean(totalSnow_cm, na.rm=TRUE))


plot(x=meaned_wd$snow,y=meaned_wd$num_trips,xlab = "Snow",sub = "Snow vs no. of trip",ylab = "Number of trips(per day)")
par(mfrow = c(2, 1))
meaned_wd = start_df %>%
    group_by(k_date) %>%
    dplyr::summarize(num_trips = mean(count, na.rm=TRUE),rain = mean(precipMM, na.rm=TRUE))

plot(x=meaned_wd$rain,y=meaned_wd$num_trips,xlab = "Rain",sub = "Rain vs no. of trip",ylab = "Number of trips(per day)")
library(corrplot)
str(start_df)
test = start_df[c("uvIndex","tempC","totalSnow_cm","mintempC","maxtempC","windSpeedKmph","precipMM","humidity","visibilityKM","cloudcover","FeelsLikeC")]
test = scale(test)
corrplot(test)
```

# plot top stations based on station #
```{r}
# plot top stations based on START station
top_stations = aggregate(x= start_df$count,
	          by= list(start_df$start_station_name),
	          FUN=sum)
top_stations = top_stations[order(top_stations$x,decreasing=T),]
top_stations = top_stations[1:20,]

ggplot(data=top_stations, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))

# plot top stations based on END station
top_stations = aggregate(x= end_df$count,
	          by= list(end_df$end_station_name),
	          FUN=sum)
top_stations = top_stations[order(top_stations$x,decreasing=T),]
top_stations = top_stations[1:20,]

ggplot(data=top_stations, aes(x=Group.1, y=x)) +
  geom_bar(stat="identity")+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5))+
  labs(x="station names",
       y="Number of trips ended at")
```


```{r}
rm(list = ls())
```
